### 2022-07-23 latex/Chapters/Code.tex
update

\section{Code organisation}

@@

\subsection{rubiks.core}
This submodule contains bases classes that make the code base easier to use, debug, and extend. It contains three base classes:
\begin{itemize}
\item Loggable: a wrapper around Python's logger which automatically picks up classes' names at init and format things (dict, series and dataframes in particular) in a nicer way.
\item Parsable: a wrapper around ArgumentParser, which allows to construct objects in the project from command line, to define dependencies between object's configurations and to help a bit with typing of configs. The end result is that you can pretty much pass **kw\_args everywhere and it just works.
\item Factory: a typical factory pattern. Concrete factories can just define what widget they produce and the factory will help construct them from **kw\_args (or command line, since Factory inherits from Parsable)
\end{itemize}

\subsection{rubiks.puzzle}
This submodule contains:
\begin{itemize}
\item Puzzle: a Factory of puzzles. It defines states and actions in the abstract, and provides useful functions to apply moves, shuffle, generate training sets, tell if a state is the goal, etc. Puzzle can manufacture the two following types of puzzles:
\item SlidingPuzzle. Implements the states and moves of the sliding puzzle.
\item RubiksCube. Implements the states and moves of the Rubik's cube.
In addition, in contains a Puzzled base class which most below inherit from. That allow e.g. heuristics, search algorithms, solvers and learners to know what puzzle and dimension they operate on, without having to reimplement these basic facts in each of them.
\end{itemize}

\subsection{rubiks.search}
This modules contains graph search strategies. I have actually reused the code I implemented for one of the AIPnT assignments. It contains the following classes:
\begin{itemize}
\item Node: which contains the state of a graph, as well as link to the previous (parent) state, action that leads from the latter to the former and the cost of the path so far.
\item SearchStrategy, a Factory class which can instantiate the following three types of search strategies to find a path to a goal:
\item BreadthFirstSearch, which is obviously an optimal strategy, but not particularly efficient.
\item DepthFirstSearch, which is not an optimal strategy, and also generally not particularly efficient.
\item AStar, which is optimal, and as efficient as the heuristic it makes use of is.
\end{itemize}

\subsection{rubiks.heuristics}
\label{HSS}
This module contains base class Heuristic, also a Factory. Heuristic can instantiate the following heuristics, which we can use in the AStar strategy from the previous section:
\begin{itemize}
\item Manhattan: at current time of writing, this is specific to the SP and will be discussed in more details in \ref{S33},
\item PerfectHeuristic: this reads from a data base the optimal costs, pre-computed by the PerfectLearner (see below \ref{PLcode})
\item DeepLearningHeuristic: this uses a network which has been trained using DRL by the DeepReinforcementLearner (see below \ref{DRLcode})
\end{itemize}

\subsection{rubiks.deeplearning}
This module is a wrapper around Pytorch. It contains:
\begin{itemize}
\item DeepLearning: a Puzzled Loggable Factory that can instantiate some configurable deep networks, and provide the necessary glue with the rest of the code base so that puzzles be seemlessly passed to the networks and trained on.
\item FullyConnected: wrapper around a Pytorch fully connected network, with configurable layers and size.
\item Convolutional: tdb
\end{itemize}

\subsection{rubiks.learners}
\label{PLcode}
\label{DRLcode}
This module implements learners, which learn something from a puzzle, store what they learnt, and can display interesting things about what they learnt.

@@

\subsection{rubiks.solvers}
This module implements solvers, which solve puzzles. The base class Solver is a Factory of solvers, and in addition to being able to instantiating the following types of solvers, can run different solvers through a similal sequences of random puzzles (for various increasing degrees of difficulty (shuffling), and/or perfectly shuffled ones) and display a comparison of how they perform in a number of metrics.

@@

\subsection{rubiks.scripts}

@@

%-----------------------------------
%	SECTION 2
%-----------------------------------
\section{Learners}

Here I show how to run some examples... TBC

%-----------------------------------
%	SUBSECTION 2.1
%-----------------------------------
\subsection{Perfect Learner}
\label{PLSS}

%-----------------------------------
%	SUBSECTION 2.2
%-----------------------------------
\subsection{Deep Learner}
\label{DLSS}
blabla

%-----------------------------------
%	SUBSECTION 2.3
%-----------------------------------
\subsection{Deep Reinforcement Learner}
\label{DRLSS}

blabla

%-----------------------------------
%	SECTION 3
%-----------------------------------

\section{Solvers}

%-----------------------------------
%	SUBSECTION 3.1
%-----------------------------------
\subsection{BFS}
\label{BFSSS}

%-----------------------------------
%	SUBSECTION 3.2
%-----------------------------------
\subsection{DFS}
\label{DFSSS}

%-----------------------------------
%	SUBSECTION 3.3
%-----------------------------------
\subsection{A*}
\label{ASSS}

\textbf{manhattan heuristic}

\textbf{perfect heuristic}

\textbf{deep learning heuristic}

\textbf{deep reinforcement learning heuristic}

>>>>

%\section{Code organisation}

@@

\section{rubiks.core}
This submodule contains base classes that make the code base easier to use, debug, and extend. It contains the following:
\begin{itemize}
\item Loggable: a wrapper around Python's logger which automatically picks up classes' names at init and format things (dict, series and dataframes in particular) in a nicer way.
\item Parsable: a wrapper around ArgumentParser, which allows to construct objects in the project from command line, to define dependencies between object's configurations and to help a bit with typing of configs. The end result is that you can pretty much pass **kw\_args everywhere and it just works.
\item Factory: a typical factory pattern. Concrete factories can just define what widget they produce and the factory will help construct them from **kw\_args (or command line, since Factory inherits from Parsable)
\end{itemize}

\section{rubiks.puzzle}
This submodule contains:
\begin{itemize}
\item Puzzle: a Factory of puzzles. It defines states and actions in the abstract, and provides useful functions to apply moves, shuffle, generate training sets, tell if a state is the goal, etc. Puzzle can manufacture the two following types of puzzles:
\item SlidingPuzzle. Implements the states and moves of the sliding puzzle.
\item RubiksCube. Implements the states and moves of the Rubik's cube.
In addition, in contains a Puzzled base class which most below inherit from. That allow e.g. heuristics, search algorithms, solvers and learners to know what puzzle and dimension they operate on, without having to reimplement these basic facts in each of them.
\end{itemize}

\section{rubiks.search}
This modules contains graph search strategies. I have actually reused the code I implemented for one of the AIPnT assignments. It contains the following classes:
\begin{itemize}
\item Node: which contains the state of a graph, as well as link to the previous (parent) state, action that leads from the latter to the former and the cost of the path so far.
\item SearchStrategy, a Factory class which can instantiate the following three types of search strategies to find a path to a goal:
\item BreadthFirstSearch, which is obviously an optimal strategy, but not particularly efficient.
\item DepthFirstSearch, which is not an optimal strategy, and also generally not particularly efficient.
\item AStar, which is optimal, and as efficient as the heuristic it makes use of is.
\end{itemize}

\section{rubiks.heuristics}
\label{HSS}
This module contains base class Heuristic, also a Factory. Heuristic can instantiate the following heuristics, which we can use in the AStar strategy from the previous section:
\begin{itemize}
\item Manhattan: at current time of writing, this is specific to the SP and will be discussed in more details in \ref{S33},
\item PerfectHeuristic: this reads from a data base the optimal costs, pre-computed by the PerfectLearner (see below \ref{PLcode})
\item DeepLearningHeuristic: this uses a network which has been trained using DRL by the DeepReinforcementLearner (see below \ref{DRLcode})
\end{itemize}

\section{rubiks.deeplearning}
This module is a wrapper around Pytorch. It contains:
\begin{itemize}
\item DeepLearning: a Puzzled Loggable Factory that can instantiate some configurable deep networks, and provide the necessary glue with the rest of the code base so that puzzles be seemlessly passed to the networks and trained on.
\item FullyConnected: wrapper around a Pytorch fully connected network, with configurable layers and size.
\item Convolutional: tdb
\end{itemize}

\section{rubiks.learners}
\label{PLcode}
\label{DRLcode}
This module implements learners, which learn something from a puzzle, store what they learnt, and can display interesting things about what they learnt.

@@

\section{rubiks.solvers}
This module implements solvers, which solve puzzles. The base class Solver is a Factory of solvers, and in addition to being able to instantiating the following types of solvers, can run different solvers through a similal sequences of random puzzles (for various increasing degrees of difficulty (shuffling), and/or perfectly shuffled ones) and display a comparison of how they perform in a number of metrics.

@@

\section{rubiks.scripts}

@@

### 2022-07-23 latex/Chapters/Puzzles.tex
update

The first puzzle I will focus on is the sliding puzzle (see \cite{SlidingPuzzleWiki}).The 15-puzzle seems to have been invented in the late 19th century by Noyes Chapman (see \cite{SlidingPuzzleWolfram}), who applied for a patent in 1980. In 1879, a couple of interesting notes (\cite{Johnson1879}), published in the American Journal of Mathematics proved that exactly half of the $16!$ possible ways of setting up the 16 tiles on the board lead to solvable puzzles. A more modern proof can be found in \cite{Archer1999}.
\\
\\
Since then, several variations of the 15-puzzle have become popular, such as the 24-puzzle. A rather contrived but interesting one is the coiled 15-puzzle (\cite{Coiled15Puzzle}), where the bottom-right and the top-left compartments are adjascent; the additional move that this allows renders all $16!$ configurations solvable.

@@

\begin{fifteen}
\setrow{4}{1,6,2,3}
\setrow{3}{5,10,7,4}
\setrow{2}{9,15,14,11}
\setrow{1}{13,12,,8}
\end{fifteen}
\\
\\
we have p(c) = (1, 6, 2, 3, 4, 7, 10, 5, 9, 15, 14, 11, 8, 12, 13).
\\
It is easy to see that the parity of p(c) cannot change by a legal move of the puzzle. Indeed, p(c) is clearly invariant by lateral move of a tile, so its parity is invariant too. A vertical move of a tile will displace a number in p(c) by an even number of positions right or left. For instance, moving tile 14 into the empty compartment above it results in a new configuration $c_{2}$:

\begin{fifteen}
\setrow{4}{1,6,2,3}
\setrow{3}{5,10,7,4}
\setrow{2}{9,15,,11}
\setrow{1}{13,12,14,8}
\end{fifteen}
\\
\\
with $p(c_{2})$ = (1, 6, 2, 3, 4, 7, 10, 5, 9, 15, 11, 8, 14, 12, 13), which is equivalent to moving 14 by 2 positions on the right. This obviously cannot change the parity since exactly 2 pairs of numbers are now in a different order, that is (14, 11) and (14, 8) now appear in the respective opposite orders as (11, 14) and (8, 14).
\\
\\
This is the crux of the proof of the well known necessary condition (even parity of p(c)) for a configuration c to be solvable (see part I of \cite{Johnson1879}).
\\
\\
In the case of the coiled puzzle, we can clearly solve all configurations of even parity, since all the legal moves of the normal puzzle are allowed. In addition, we can for instance transition between the following 2 configurations, which clearly have respectively even and odd parities:

\begin{fifteen}
\setrow{4}{1,2,3,4}
\setrow{3}{5,6,7,8}
\setrow{2}{9,10,11,12}
\setrow{1}{13,14,15,}
\end{fifteen}
%
\begin{fifteen}
\setrow{4}{,2,3,4}
\setrow{3}{5,6,7,8}
\setrow{2}{9,10,11,12}
\setrow{1}{13,14,15,1}
\end{fifteen}
\\
\\
Since it is possible to reach an odd parity configuration, and by symmetry arguments, we conclude we can solve all $16!$ configurations.

>>>>

The first puzzle I will focus on is the sliding puzzle (see \cite{SlidingPuzzleWiki}).The 15-puzzle seems to have been invented in the late 19th century by Noyes Chapman (see \cite{SlidingPuzzleWolfram}), who applied in 1880 for a patent on what was then called "Block Solitaire Puzzle". In 1879, a couple of interesting notes (\cite{Johnson1879}), published in the American Journal of Mathematics proved that exactly half of the $16!$ possible ways of setting up the 16 tiles on the board lead to solvable puzzles. A more modern proof can be found in \cite{Archer1999}.
\\
\\
Since then, several variations of the 15-puzzle have become popular, such as the 24-puzzle. A rather contrived but interesting one is the coiled 15-puzzle (\cite{Coiled15Puzzle}), where the bottom-right and the top-left compartments are adjascent; the additional move that this allows renders all $16!$ configurations solvable.

@@

\begin{fifteen}
\setrow{4}{1,6,2,3}
\setrow{3}{5,10,7,4}
\setrow{2}{9,15,\red14\black,\blue11}
\setrow{1}{13,12,,\blue 8}
\end{fifteen}
\\
\\
we have p(c) = (1, 6, 2, 3, 4, 7, 10, 5, 9, 15, \red 14 \black, \blue 11, 8\black, 12, 13).
\\
It is easy to see that the parity of p(c) cannot change by a legal move of the puzzle. Indeed, p(c) is clearly invariant by lateral move of a tile, so its parity is invariant too. A vertical move of a tile will displace a number in p(c) by an even number of positions right or left. For instance, moving tile 14 into the empty compartment below it results in a new configuration $c_{2}$:

\begin{fifteen}
\setrow{4}{1,6,2,3}
\setrow{3}{5,10,7,4}
\setrow{2}{9,15,,\blue11}
\setrow{1}{13,12,\red14\black,\blue8}
\end{fifteen}
\\
\\
with $p(c_{2})$ = (1, 6, 2, 3, 4, 7, 10, 5, 9, 15, \blue11, 8\black, \red14\black, 12, 13), which is equivalent to moving 14 by 2 positions on the right. This obviously cannot change the parity since exactly 2 pairs of numbers are now in a different order, that is (14, 11) and (14, 8) now appear in the respective opposite orders as (11, 14) and (8, 14).
\\
\\
This is the crux of the proof of the well known necessary condition (even parity of p(c)) for a configuration c to be solvable (see part I of \cite{Johnson1879}).
\\
\\
In the case of the coiled puzzle, we can clearly solve all configurations of even parity, since all the legal moves of the normal puzzle are allowed. In addition, we can for instance transition between the following 2 configurations, which clearly have respectively even and odd parities:

\begin{fifteen}
\centering
\setrow{4}{\red1\black,2,3,4}
\setrow{3}{5,6,7,8}
\setrow{2}{9,10,11,12}
\setrow{1}{13,14,15,}
\end{fifteen}
%
\begin{fifteen}
\setrow{4}{,2,3,4}
\setrow{3}{5,6,7,8}
\setrow{2}{9,10,11,12}
\setrow{1}{13,14,15,\red1\black}
\end{fifteen}
\\
\\
Since it is possible to reach an odd parity configuration, we conclude by invoking symmetry arguments that we can solve all $16!$ configurations.

### 2022-07-23 latex/MScAIFrancoisBerrier.tex
update

\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
\usepackage{mathpazo} % Use the Palatino font by default
\usepackage[T5,T1]{fontenc}
\usepackage{verbatim}
\usepackage[backend=bibtex,style=authoryear,natbib=true]{biblatex} % Use the bibtex backend with the authoryear citation style (which resembles APA)
\usepackage{float}
\usepackage{lscape}
\usepackage{pdflscape}
\usepackage{rotating}
\usepackage[paper=portrait,pagesize]{typearea}
\usepackage{logicpuzzle}
\newenvironment{fifteen}[1][]{%
\begin{logicpuzzle}[rows=4,columns=4,#1]
\begin{puzzleforeground}
\framepuzzle
\end{puzzleforeground}
}{\end{logicpuzzle}}

@@

\begin{abstract}
\addchaptertocentry{\abstractname} % Add the abstract to the table of contents
\textbf{Motivation}
\\
Reinforcement Learning (\textbf{RL}), exposed with brilliant clarity in the Sutton book (\cite{Sutton1998}), has until recently known less success than we might have hoped for. Its framework is very appealing and intuitive. In particular, the mathematical beauty of Value iteration and Q iteration (\cite{WatkinsThesis}) for discrete state and action spaces, blindly iterating from \textit{any} initial value, is quite profound. Sadly, it had until recently proven hard to achieve practical success with these methods.
\\
Inspired however by the seminal success of Deep Mind's team in using Deep Reinforcement Learning (\textbf{DRL}) to play Atari games (\cite{Mnih2013})  and to master the game of Go (\cite{AlphaGo}), researchers have in recent years made a lot of progress towards designing algorithms capable of learning and solving, \textit{without human knowledge}, the Rubiks Cube (\textbf{RC}) - as well as similar single player puzzles - by using Deep Q-Learning (\textbf{DQR})  (\cite{DBLP:journals/corr/abs-1805-07470}) or search and (\textbf{DRL}) value iteration (\cite{https://doi.org/10.48550/arxiv.1805.07470}).
\\
In this project, I will attempt to implement a variety of solvers combining $A^{*}$ search and heuristics, some of which will be handcrafted, others which I will train on randomly generated sequences of puzzles using (\textbf{DL}) or (\textbf{DRL}), to solve the 15-puzzle (and variations of different dimensions), as well as the Rubik's cube.
\\
\\
\textbf{Organisation of this thesis}
\\
In the first chapter, I will quickly describe what I hope to get out of this project, both in terms of personal learning, as well as in terms of tangible results (solving some puzzles!). In chapter 2, I will do a quick recap of the different methods that I will use in the project. Chapter 3 will be dedicated to discussing the mathematics of the sliding puzzle (\textbf{SP}) and the (\textbf{RC}). I might throw a few random (but hopefully interesting) observations in there and give some references for the keen reader. I will then give an overview in chapter 4 of the code base I have developed - and put in the open on my github page (\cite{FB}) - to complete this project. Finally, in chapters 5 and 6, I will present all my various results on respecively the sliding puzzle and the Rubiks' cube.

@@

\include{Chapters/Objectives}
\include{Chapters/DRLSearch}
\include{Chapters/Puzzles}
\include{Chapters/Code}
\include{Chapters/SlidingPuzzle}
\include{Chapters/Rubiks}

>>>>

\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
\usepackage{mathpazo} % Use the Palatino font by default
\usepackage[T5,T1]{fontenc}
\usepackage{verbatim}
\usepackage[backend=bibtex,style=authoryear,natbib=true]{biblatex} % Use the bibtex backend with the authoryear citation style (which resembles APA)
\usepackage{float}
\usepackage{graphbox}
\usepackage{lscape}
\usepackage{pdflscape}
\usepackage{rotating}
\usepackage[paper=portrait,pagesize]{typearea}
\usepackage{logicpuzzle}
\usepackage{geometry}
\newenvironment{three}[1][]{%
\begin{logicpuzzle}[rows=2,columns=2,#1]
\begin{puzzleforeground}
\framepuzzle
\end{puzzleforeground}
}{\end{logicpuzzle}}

\newenvironment{five}[1][]{%
\begin{logicpuzzle}[rows=2,columns=3,#1]
\begin{puzzleforeground}
\framepuzzle
\end{puzzleforeground}
}{\end{logicpuzzle}}

\newenvironment{seven}[1][]{%
\begin{logicpuzzle}[rows=2,columns=4,#1]
\begin{puzzleforeground}
\framepuzzle
\end{puzzleforeground}
}{\end{logicpuzzle}}

\newenvironment{nine}[1][]{%
\begin{logicpuzzle}[rows=2,columns=5,#1]
\begin{puzzleforeground}
\framepuzzle
\end{puzzleforeground}
}{\end{logicpuzzle}}

\newenvironment{eight}[1][]{%
\begin{logicpuzzle}[rows=3,columns=3,#1]
\begin{puzzleforeground}
\framepuzzle
\end{puzzleforeground}
}{\end{logicpuzzle}}

\newenvironment{fifteen}[1][]{%
\begin{logicpuzzle}[rows=4,columns=4,#1]
\begin{puzzleforeground}
\framepuzzle
\end{puzzleforeground}
}{\end{logicpuzzle}}

\newenvironment{twenty}[1][]{%
\begin{logicpuzzle}[rows=4,columns=5,#1]
\begin{puzzleforeground}
\framepuzzle
\end{puzzleforeground}
}{\end{logicpuzzle}}

\newenvironment{twentyfour}[1][]{%
\begin{logicpuzzle}[rows=5,columns=5,#1]
\begin{puzzleforeground}
\framepuzzle
\end{puzzleforeground}
}{\end{logicpuzzle}}

\newenvironment{thityfive}[1][]{%
\begin{logicpuzzle}[rows=6,columns=6,#1]
\begin{puzzleforeground}
\framepuzzle
\end{puzzleforeground}
}{\end{logicpuzzle}}

\newcommand{\red}{\color{red}}
\newcommand{\black}{\color{black}}
\newcommand{\blue}{\color{blue}}
\newcommand{\green}{\color{green}}

\newenvironment{titlemize}[1]{%
  \paragraph{#1}
  \begin{enumerate}}
  {\end{enumerate}}

@@

\begin{abstract}
\addchaptertocentry{\abstractname} % Add the abstract to the table of contents
\textbf{Motivation}
\\
Reinforcement Learning (\textbf{RL}), exposed with brilliant clarity in the Sutton book (\cite{Sutton1998}), has until recently known less success than we might have hoped for. Its framework is very appealing and intuitive. In particular, the mathematical beauty of Value iteration and Q iteration (\cite{WatkinsThesis}) for discrete state and action spaces, blindly iterating from \textit{any} initial value, is quite profound. Sadly, it had until recently proven hard to achieve practical success with these methods.
\\
Inspired however by the seminal success of Deep Mind's team in using Deep Reinforcement Learning (\textbf{DRL}) to play Atari games (\cite{Mnih2013})  and to master the game of Go (\cite{AlphaGo}), researchers have in recent years made a lot of progress towards designing algorithms capable of learning and solving, \textit{without human knowledge}, the Rubiks Cube (\textbf{RC}) - as well as similar single player puzzles - by using Deep Q-Learning (\textbf{DQR})  (\cite{DBLP:journals/corr/abs-1805-07470}) or search and (\textbf{DRL}) value iteration (\cite{https://doi.org/10.48550/arxiv.1805.07470}).
\\
In this project, I will attempt to implement a variety of solvers combining $A^{*}$ search and heuristics, some of which will be handcrafted, others which I will train on randomly generated sequences of puzzles using (\textbf{DL}) or (\textbf{DRL}), to solve the 15-puzzle (and variations of different dimensions), as well as the Rubik's cube.
\\
\\
\textbf{Organisation of this thesis}
\\
In the first chapter, I will quickly describe what I hope to get out of this project, both in terms of personal learning, as well as in terms of tangible results (solving some puzzles!). In chapter 2, I will do a quick recap of the different methods that I will use in the project. Chapter 3 will be dedicated to discussing the mathematics of the sliding puzzle (\textbf{SP}) and the (\textbf{RC}). I might throw a few random (but hopefully interesting) observations in there and give some references for the keen reader. I will then give an overview in chapter 4 of the code base I have developed - and put in the open on my github page (\cite{FB}) - to complete this project before detailing a few examples in chapter 5. Finally, in chapters 6 and 7, I will present all my various results on respectively the sliding puzzle and the Rubiks' cube.

@@

\include{Chapters/Objectives}
\include{Chapters/DRLSearch}
\include{Chapters/Puzzles}
\include{Chapters/Code}
\include{Chapters/Examples}
\include{Chapters/SlidingPuzzle}
\include{Chapters/Rubiks}
